Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias

### 2021-06-06 04_reports/manuscript.Rmd
changed the time-varying approach including tv-covs to the t2cancer proxy

---
title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
author: "L. Paloma Rojas-Saunero"
date: ""
output: [html_document]
editor_options: 
  chunk_output_type: console
bibliography: cancer_dementia.bib
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(kableExtra)
source(here::here("02_R", "03_descriptives.R"))
source(here::here("02_R", "05_plots.R"))
source(here::here("02_R", "05b_plot_causes_death.R"))
# table_results <- rio::import(here::here("02_R", "table_results.csv"))
# table_results_bounds <- rio::import(here::here("02_R", "table_results_bounds.csv"))
# models <- read_csv(here::here("04_reports", "models.csv")) %>% select(-6)
```

@@

Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward. Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results[@driverbiogeront2014;@ganguli2015]. However, understanding these or other sources of bias first requires making explicit the causal question. Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017]. 
To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on a specific question conceptualizing the  Pin1 enzyme as the target of intervention. Previous animal studies have shown that Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition[@driverpin12015;@angelucci2017;@li2021]. If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_

@@

Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it. Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020]. Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a time-varying process. We depict this feature in Figure 2, where $P^*_{t-1}$ and $P^*_t$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t-1}$ and $P_t$ respectively. Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.

@@

This problem is avoided by recognizing the time-varying nature of cancer diagnosis. Several studies have considered cancer diagnosis as a time-dependent exposure. In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017]. The price we pay with this approximation is that implicitly, this means that Pin1 would overexpress at the time of cancer diagnosis, which is biologically implausible. Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up. Thus we included death prior to cancer diagnosis as $D_t$ and an arrow between $D_t$ and $P_t$ that represents a deterministic association such as that $P_t$ is only observed if $D_t$ is zero in Figure 3. In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as COPD.

@@

We now move to the challenge of death as a competing event for ADRD, represented in Figure 5. For interpretability we exclude the time varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$. Given that some participants may die prior to having a cancer diagnosis over follow-up, we can only have ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis. In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies prior to ADRD diagnosis, death prevents observing ADRD at future time-points. Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions. We also include an arrow between $L$ and $D_{t+1}$ following the argument discussed previously.

In the scenario where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020]. However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$. In this setting, a total effect of $P_t$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].

To isolate the direct effect of $P_t$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives of causal questions we can ask. We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020]. Alternatively, we can define an scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$. Previous studies have defined death as a censoring event [ref], although failed to explicitly define how to relax the independent censoring assumption. Moreover, adjusting for confounders for dementia may represent partially the adjustment for the competing event of death [@young2020], adjusting for baseline covariates may not be sufficient.

@@

To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 6. Even so, this is a simplified version since we omitted the time varying nature of all nodes in the graph and other sources of measurement error, and thus are ignoring how feedback loops between them further complicate interpretability and identificability.

@@

## 4. Application to the Rotterdam Study

@@

### 4.1. Methods

We illustrate the association between cancer and dementia diagnosis under two scenarios. For the first scenario (Model A) we defined cancer proxy as _"cancer ever vs. never"_.This scenario mimics the definition used in previous studies where cancer was defined as a time-independent exposure[@driver2012; @nudelman2014]. We continue with a second scenario (Model B) where cancer diagnosis is a time-varying exposure. The definition represents the setting where all participants are free of cancer at inclusion and over time some participants will have a cancer diagnosis.

To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile. Weights for model A were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status. In contrast, for Model B weights recreate a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000]. Time varying covariates included: ...
Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.

To estimate the controlled direct effect of Pin1 in the risk of ADRD, that is in the setting where death could have been prevented, we fit inverse probability censoring weights for death.

in time-varying settings we compared the complement of a weighted Kaplan-Meier survival estimator in participants with incident cancer vs. no incident of cancer with time indexed in years. The weights in this case are time-varying by follow-up year, defined as a product of the time-fixed IPT weights above and a year-specific inverse probability of censoring (IPC) by death weights.  For an individual still alive in year t, the time t IPC weight is the product of the inverse probability of surviving in each year prior to t, conditional on measured common causes of death and dementia (that is, variables such as C in Figure 4). For an individual who has died by time t, the year t IPC weight is zero.  We estimated survival probabilities using a logistic regression model for death as a function of baseline and time-varying covariates. Baseline covariates included age at study entry, sex, apoe4 status, educational attainment and the time-varying covariates smoking status, systolic blood pressure, BMI and prevalent and incident comorbidities such as: cancer, heart disease, stroke and diabetes. We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{15}$ and $Y_{20}$) for an illustrative purpose.

Estimates of the controlled direct effect at 20 years of follow-up are presented as risk differences (RD), risk ratios (RR) and hazard ratios (HR). We note that hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case. We present them for comparison against risk ratios.

Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound). The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan Meier estimator for the combined outcome of dementia or death.

We additionally present the causes of death for participants who developed cancer and for participants free of cancer diagnosis over follow-up. This information provides insights about the association between cancer diagnosis and death and about the potential misclassification of individuals with no cancer diagnosis (cancer as a cause of death).

All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples. All analysis were performed using R, code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.

>>>>

---
title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
author: ""
date: ""
output: [pdf_document]
always_allow_html: true
editor_options: 
  chunk_output_type: console
bibliography: cancer_dementia.bib
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(flextable)
source(here::here("02_R", "03_descriptives.R"))
source(here::here("02_R", "table_2.R")) 
load(here::here("03_figs", "plots.RData")) 
source(here::here("02_R", "05_plots.R"))
```

@@

Nevertheless, inferring any treatment or mechanistic effects from the observed cancer-ADRD inverse association is not straightforward. Researchers have raised concerns related to the competing event of death, unmeasured confounding, and ascertainment error that could explain these results[@driverbiogeront2014;@ganguli2015]. However, understanding these or other sources of bias first requires making explicit the causal question. Moreover, making explicit the causal question is one step toward tying a research study to a question that is relevant to decision-making[@didelez2016;@labrecque2017].

To illustrate the complexities of inferring hypothetical or available treatments’ effects on ADRD from the observed cancer-ADRD association, we focus on a specific question conceptualizing the  Pin1 enzyme as the target of intervention. Previous animal studies have shown that Pin1 enzyme over-expression promotes tumorigenesis, while its down-regulation is attributed to mechanisms that contribute to neurodegeneration and amyloid deposition[@driverpin12015;@angelucci2017;@li2021]. If we one day could develop a drug that increases Pin1 expression specifically in brain tissue in hopes of preventing dementia, we could pose the question as: _What is the effect of this Pin1-targeting drug on the risk of ADRD over time compared to standard treatments?_

@@

Currently, Pin1 expression is not an available biomarker for population-based research, so at best we can only rely on a proxy for it. Because Pin1 over-expression is present in tumors, and tumors are only measured through diagnosis, previous studies have considered Pin1 over-expression as the explanation behind the inverse association between cancer diagnosis and ADRD, though Pin1 was not explicitly part of the research question[@ospina2020;@driver2012;@musicco2013;@freedman2016;@bowles2017;@frain2017;@schmidt2017;@sun2020;@ording2020]. Unlike measuring Pin1 at the same time to all participants (thought this would not necessarily mean this would be the ideal time to measure it, we discuss this point further in the discussion section), cancer diagnosis is a collected over time. We depict this feature in Figure 2, where $P^*_{t-1}$ and $P^*_t$ represent _cancer diagnosis_ over time, the measured proxy of $P_{t-1}$ and $P_t$ respectively. Although this means we would measure the association between cancer diagnosis over time and ADRD in the observed data, we are assuming that the captured effect is only through the pathway that involves Pin1 expression over time.

@@

This problem is avoided by recognizing the time-varying nature of cancer diagnosis. Several studies have considered cancer diagnosis as a time-dependent exposure. In this way, the time prior to cancer diagnosis is allocated to the non-exposed arm, and the time after cancer diagnosis to the exposed arm [@white2013; @hanson2016; @bowles2017]. The price we pay with this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis, which is biologically implausible. Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up. Thus we included death prior to cancer diagnosis as $D_t$ and an arrow between $D_t$ and $P_t$ that represents a deterministic association such as that $P_t$ is only observed if $D_t$ is zero in Figure 3. In addition, we added an arrow between $L$ and $D_t$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as COPD. Although $L$ is time-varying in nature, we only depict $L$ at one time-point for readability.

@@

We now move to the challenge of death as a competing event for ADRD, represented in Figure 4. For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$. Given that some participants may die prior to having a cancer diagnosis over follow-up, we can only have ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis. In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia. Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions. We also include an arrow between $L$ and $D_{t+1}$ following the argument discussed previously.

In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020]. However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$. In this setting, a total effect of $P_t$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].

To isolate the direct effect of $P_t$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives of causal questions we can ask. We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020]. Alternatively, we can define an scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$. Previous studies have defined death as a censoring event [], although failed to explicitly define how to comply with the independent censoring assumption. Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) may represent partially the adjustment for the competing event of death [@geskus], to block the pathway mediated by $C$ we need to consider covariates that change over time and that affect both nodes, such as cardiometabolic conditions.

@@

To summarize, the complexity of the causal structure that describes the effect of Pin1 through the proxy of cancer diagnosis in the risk of ADRD illustrate the potential sources of bias, as observed in Figure 5. Even so, this is a simplified version since we omitted other sources of measurement error [@whatif2020] and the time-varying nature of all nodes and feedback loops between them which would further complicate interpretability and identificability.

@@

## 3. Application to the Rotterdam Study

@@

### 3.1. Methods

We illustrate the association between cancer and dementia diagnosis under two scenarios. Scenario A replicates the setting that defines cancer proxy as _"cancer ever vs. never"_ [@driver2012]. Scenario B defines cancer diagnosis is a time-varying exposure. To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile. In scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status. In contrast, for scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on time-varying covariate history. This recreates a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000]. Time varying covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident comorbidities such as: hypertension and diabetes. Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.

Inverse probability censoring weights for death were calculated to relax the independent censoring assumption. In scenario A, weights represent the inverse of the probability of not dying conditional on baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort. For individuals who died, their censoring weight is zero [@whatif2020]. In scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia. For an individual who has died by time _t_, the year _t_ censoring weight is zero. Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.

To estimate the controlled direct effect of Pin1 in the risk of ADRD, we compared the complement of a weighted Kaplan-Meier survival estimator for participants who developed cancer versus those who did not, with time indexed in years. The weights are time-varying by follow-up year, defined as a product of the year-specific inverse probability of treatment weights and the inverse probability of censoring by death weights. Estimates of the controlled direct effect at 20 years of follow-up are presented as risk differences (RD) and risk ratios (RR). All 95% confidence intervals were calculated using percentile-based bootstrapping based on 500 bootstrap samples.

For illustrative and comparative purposes we calculated hazard ratios (HR). Hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case [@young2020]. We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{t+1}$ and $Y_{t+1}$).

Since the conditional independent censoring assumption is untestable, we compute Peterson upper and lower bounds [@peterson1976] to represent: 1) the extreme scenario of independence, that refers to an scenario were those who died would never develop dementia (lower bound) and 2) complete dependency, that refers to an scenario where those who died would have a dementia prior to death (upper bound). The lower bound is calculated with the Aalen-Johansen estimator treating death as a competing event, and the upper bound is calculated with the Kaplan Meier estimator for the combined outcome of dementia or death.

All analysis were performed using R, code is provided in supplementary material and available in https://github.com/palolili23/2021_cancer_dementia.

### 3.2. Results

Participants had a mean age of `r mean_age`, and `r women_prop` were women. Further details on participants are presented in Table 1. Over follow-up, `r yes_cancer` developed cancer and `r no_cancer` remain free of cancer diagnosis, the median age of cancer diagnosis was `r cancer_time`. From the total sample, `r dem_total` had dementia over follow-up and median time to dementia was `r dem_time`. Among participants with incident cancer, `r cancer_dem` had dementia diagnosis and `r cancer_death` died over follow-up, `r cancer_alive` remain alive at 20 years since study entry. In contrast, among participants free of cancer diagnosis over follow-up, `r no_cancer_dem` were diagnosed with dementia and `r no_cancer_death` died over follow-up, `r no_cancer_alive` were alive at the end of follow-up. Figure x. shows the magnitude of the proportion of people who died over time.

Results for all scenarios are present in Table 2. Had we defined the proxy for Pin1 as _ever cancer_, and relying on death as independent censoring event (unconditional), we observe a significant protective effect of ever having cancer in the risk of dementia [RR: `r complete_table[3,4]`, HR:  `r complete_table[3,5]`]. This effect did not change after including censoring weights for death [RR: `r complete_table[2,4]`; HR: `r complete_table[2,5]`].

In contrast, had we defined cancer diagnosis as a _time-varying_ proxy for Pin1 over-expression and had we prevented death conditional on covariates, the risk of dementia had participants had a cancer diagnosis over time is higher [RR: `r complete_table[6,4]`; HR: `r complete_table[6,5]`], though confidence intervals cross the null.

\newpage

## Tables

### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow-up.

```{r}
tableone_exp <-
  print(
    tableone,
    explain = TRUE,
    test = FALSE,
    quote = FALSE,
    noSpaces = TRUE,
    printToggle = FALSE
  )

kableExtra::kable(tableone_exp, booktabs = T) %>%
  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))

```

\newpage

### Table 2. Risk difference and risk ratio for the risk of dementia

```{r}
kableExtra::kable(complete_table, booktabs = T) %>%
  kableExtra::kable_styling(latex_options = c("striped", "hold_position"))
```

\newpage

## Figures

### Figure 1: Distribution of participants under each health status, by age over follow-up

```{r}
plot_distribution
```

\newpage
